---
title: 'Statistical Inference: Data Exploration'
author: "Eddie Warner"
output:
  html_document:
    keep_md: yes
    theme: united
  pdf_document: default
references:
- URL: https://en.wikipedia.org/wiki/Exponential_distribution
  authors:
    name: Wikipedia
  id: wiki_ed
  title: Exponential distribution
geometry: margin=1.5cm
---

## Overview:

Explore the ToothGrowth {datasets} using exploratory data analysis, confidence levels and hypothesis testing. This data set The *Effect of Vitamin C on Tooth Growth in Guinea Pigs* represents the following study.  

### Study Discription
"The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid)."

```{r load_library, echo=FALSE, message=FALSE}
library(ggplot2)
library(plyr)
library(dplyr)
library(data.table)
library(reshape2)
```

## Data 
Load and examine data. Data consists of three columns, interestingly in comparison to ChickWeight *data(ChickWeight)* there is no identifying column for individual Guinea Pigs. So the assumption must be that each measure was taken with a different subject (pig). 

```{r load_data, echo=TRUE}

data(ToothGrowth)
summary(ToothGrowth)
str(ToothGrowth)

```

Create a factor based on dose for easier grouping and plotting.

```{r factor, echo=TRUE}
ToothGrowth$dose <- factor(ToothGrowth$dose, levels = c(0.5,1.0,2.0), labels=c("0.5 mg", "1.0 mg", "2.0 mg"),)

```

## Plots
Plot each dose and delivery method against tooth length. Remove outlaying observations

```{r box_plot, echo=TRUE, warning=FALSE}
dose_supp <- ggplot(aes(y = len, x = dose, fill = supp), data = ToothGrowth) + geom_boxplot(outlier.shape=NA)
dose_supp <- dose_supp + labs(x = "Dose", y = "Tooth Length", title = "Tooth Lenth by Dose and Delivery")
dose_supp <- dose_supp + scale_fill_discrete(
                       name="Delivery",
                       labels=c("Orange Juice", "Ascorbic Acid"))
dose_supp

```
  
Empirically, from examining the plot, dose of vitamin c would be the strongest factor for tooth length. For lower doses of vitamin c, orange Juice is more effective than assobic acid.

## Analysis

From exploration, the delivery method of Orange Juice was better at low doses. We will confirm this observation by performing Student's t-test, testing the different delivery methods for the different dose levels. 

As noted above, the analysis assumes that each *dose ~ delivery* where carried out on a different subject (pig). As such we test with the assumption that subjects are not paired and that subjects have an unequal variance.

We will assume for our **H0** hypothesis that the delivery method are equal in effectiveness. We will use the 95% confidence interval.

### Create a wide data set
```{r wide, echo=TRUE}

wide <- ToothGrowth %>% filter(supp == "OJ", dose == "0.5 mg") %>% 
    mutate(OJ_0.5mg = len) %>% select(OJ_0.5mg)
wide <- cbind(wide, ToothGrowth %>% filter(supp == "VC", dose == "0.5 mg") 
              %>% mutate(VC_0.5mg = len) %>% select(VC_0.5mg))
wide <- cbind(wide, ToothGrowth %>% filter(supp == "OJ", dose == "1.0 mg") 
              %>% mutate(OJ_1.0mg = len) %>% select(OJ_1.0mg))
wide <- cbind(wide, ToothGrowth %>% filter(supp == "VC", dose == "1.0 mg") 
              %>% mutate(VC_1.0mg = len) %>% select(VC_1.0mg))
wide <- cbind(wide, ToothGrowth %>% filter(supp == "OJ", dose == "2.0 mg") 
              %>% mutate(OJ_2.0mg = len) %>% select(OJ_2.0mg))
wide <- cbind(wide, ToothGrowth %>% filter(supp == "VC", dose == "2.0 mg") 
              %>% mutate(VC_2.0mg = len) %>% select(VC_2.0mg))

```
### Test Dose = 0.5 mg

```{r test_0.5, echo=TRUE}
dose.0.5 <- t.test(wide['OJ_0.5mg'] - wide['VC_0.5mg'], paired = FALSE, var.equal = FALSE)
dose.0.5$conf.int
dose.0.5$p.value
```
Based on this test the **H0** hypothesis is rejected for **H1** OJ is more effective at this dose.

### Test Dose = 1.0 mg

```{r test_1.0, echo=TRUE}
dose.1.0 <- t.test(wide['OJ_1.0mg'] - wide['VC_1.0mg'], paired = FALSE, var.equal = FALSE)
dose.1.0$conf.int
dose.1.0$p.value
```
Based on this test the **H0** hypothesis is rejected for **H1** OJ is more effective at this dose.

### Test Dose = 2.0 mg

```{r test_2.0, echo=TRUE}
dose.2.0 <- t.test(wide['OJ_2.0mg'] - wide['VC_2.0mg'], paired = FALSE, var.equal = FALSE)
dose.2.0$conf.int
dose.2.0$p.value
```
Based on this test the **H0** hypothesis accepted at this dose.

## Conclusion

Through exploration of the data, Orange Juice appeared to more effective at low dose (0.5 mg, 1.0 mg). Using the Student's t-test this insight into the data was confirmed. 

